---
title: "Base R and Dplyr functions"
author: "Lyn Howard"
output: webexercises::webexercises_default
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
library(webexercises)
```

<iframe width="560" height="315" src="https://www.youtube.com/embed/JfWm-ECqXts?si=RdseoRHMFwtoO9ON" data-external="1" allowfullscreen>

</iframe>

## Transcript and code

We're going to start with recapping some base R and dplyr functions that you may have already come across. Dplyr is part of the tidyverse, so I'm loading that, and also the NHSRdatasets package, which includes the data we're going to use. To run multiple lines of code in the source pane we can select the rows we want & press Ctrl + Enter.

```{r eval = FALSE, echo=TRUE}
library(tidyverse)
library(NHSRdatasets)
```


Next we load some data for attendances at Accident & Emergency (A&E). To run a single line of code we can position the cursor anywhere in the line and use Ctrl + Enter.

```{r eval = FALSE, echo=TRUE}
data <- ae_attendances
```


In the environment panel in the top right corner, by clicking the blue arrow we can see the structure of the dataframe -- over 12,000 rows of 6 columns, and it tells us the data type of those columns. We can also click on the symbol on the right which opens the dataframe in a tab in the source pane. To get the names of the columns in a dataframe we can use either **colnames()** or **names()** function.

Since we're working with a dataframe, they both give the same output.

```{r eval = FALSE, echo=TRUE}
colnames(data)      # works on matrices and data frames
names(data)         # works on vectors and data frames
```


Using dataframe name \$ column name will return the content of that column, but especially when it's a long column, it can be more useful to just find the unique data items. Now we can see there are 274 unique entries in that column.

```{r eval = FALSE, echo=TRUE}
# using $ sign to select a column
data$type

# get list of unique data items in a variable
unique(data$org_code)
```

If you know the position of a specific entry you want to retrieve, you can use square brackets. For a range of entries using the colon. \*If you're coming to R with a knowledge of Python, note that R starts counting from 1, not zero.

```{r eval = FALSE, echo=TRUE}
# we can select a specific entry within that vector with square brackets
# to call the 4th entry we can use
data$type[4]

# we could also call the 4th to 10th entries
data$type[4:10]
```

**range()** is a really useful function for working with numeric or character data, returning the maximum and minimum values.

```{r eval = FALSE, echo=TRUE}
# get the range of a variable (useful for dates)
range(data$period)
```

**str()** function shows you the structure of the dataframe, the same as you can see in the environment, but returns it as an output down in the console. 

```{r eval = FALSE, echo=TRUE}
# see structure of data to check data types
str(data)
```

The head & tail functions allow you to see a snippet of the dataframe, by default head() returns the first 5 rows and tail() the last 5 rows.  By adding a number after the data source we can change that and return however many rows we’d like to see.

```{r eval = FALSE, echo=TRUE}
# see the first 5 rows of data
head(data)

# see the last 5 rows of data
tail(data)

# see the first 15 rows of data
head(data, 
     15)
```

Dplyr has a group of similar functions called slice_. Not only can you get a number of rows, you  can also choose a proportion.  If you want to order the dataframe by a different column, you can add **order_by = **.  

```{r eval = FALSE, echo=TRUE}
# see the first 15 rows of data 
dplyr::slice_head(data, 
      n = 15)

# see the first 15% of total rows of data 
slice_head(data,
           prop = .15)

# see the first 15% of total rows of data
# ordered by attendances
slice_max(data, 
         prop = .15, 
         order_by = attendances)
```

There’s lots more information on the various slice_ functions in the help file, which you can access using **?slice** in the console pane.


## Test your understanding

```{r, echo = FALSE, results='asis'}
# Uncomment to change widget colours:
#style_widgets(incorrect = "goldenrod", correct = "purple")
```

What is the total attendance across the 5 rows with the lowest attendances? `r fitb(15)`

`r hide("See a hint")`

Use **?slice** to look at the help file to find the function you need to get a small dataframe of the 5 rows with the lowest attendance. You can then use **sum()** to get the total of the attendances column.

`r unhide()`

```{r, echo = TRUE, eval = FALSE, webex.hide = "Click here to see an explanation"}
q1_df<- slice_min(data, n=5, order_by = breaches)

sum(q1_df$attendances) # This will give the answer = 15
```
